\section*{Acknowledgments}
This thesis has been an intriguing journey just like the entire master programme has been for me. It started with an environmental interest, observing with the persistent eyes of a satellite, analyzing in multiple dimensions with tools of Big Data, geostatistics, time series, and Machine Learning to create deeper understanding of the initial problem.\\
I am thankful for what I have learned, and the chances I was given, for the freedom and the critical and very supportive feedback, not only but especially during this thesis as well as throughout the past two years. I want to thank my supervisors Jan Verbesselt and Arie Staal for just the right amount of meetings to keep me questioning myself and their energy spent on following and guiding my thoughts and their enthusiasm for the topic. I have very much enjoyed your supervisison! I want to thank Jan Verbesselt, Ute Sass-Klaassen and Mathieu Decuyper for providing me with contacts and remarks during the development of the proposal. I want to thank Jordi Mart\'{i}nez-Vilalta, Mireia Banqué, Jordi Vayreda and David Chaparro at CREAF, Barcelona, for the reference data and their feedback on the results. I want to thank Dainius Maisili\=unas for advice on the technical questions of this research and Sytze de Bruin for feedback on statistics.\\
I am grateful to my parents and my sister Hannah for their financial and emotional support that made my studies possible in the first place. I want to thank Tim Jak and Robbert-Jan Joling for proof-reading, and finally I want to thank the thesis-room community for keeping the mood up at all times and the well-deserved Thursday-evening MGI 'borrels'.\\

\bigskip

Hartelijk dank!